---
title: "Orca Encounters"
subtitle: "Data from TidyTuesday"
author: "Daniel Martínez Scavino"
format: html
execute: 
  warning: false
  message: false
---

```{r}
#| echo: false
library(tidyverse)
library(openintro)
library(maps)

```

# Orcas are nasty fish that enjoy munching on people!

Nope! There are several things that are wrong with that picture.  Orcas are cetaceans related to dolphins that live in all of the world oceans. They are higly intelligent predators that hunt in coordinated family pods using very elaborate strategies.  Their varied diet includes fish, sharks, smaller marine mammals (e.g. see lions), and even whales. But not humans! 


```{r}
#| echo: false
orcas <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-10-15/orcas.csv')
```


```{r}
#| echo: false
library(ggplot2)
library(sf)
library(ggspatial)

# Convert coordinates to sf object
orcas_sf <- orcas %>%
  filter(!is.na(begin_longitude), !is.na(begin_latitude)) %>%
  st_as_sf(coords = c("begin_longitude", "begin_latitude"), crs = 4326)

ggplot() +
  # Adds OpenStreetMap tiles automatically bounded by your data
  annotation_map_tile(type = "osm", zoom = 9) + 
  geom_sf(data = orcas_sf, color = "red", size = 1, alpha = 0.6) +
  theme_minimal() +
  labs(title = "Orca Encounter Locations")
```


```{r}
#| echo: false
library(ggplot2)
library(sf)
library(ggspatial)

# Convert coordinates to sf object
orcas_sf <- orcas %>%
  filter(!is.na(begin_longitude), !is.na(begin_latitude)) %>%
  st_as_sf(coords = c("begin_longitude", "begin_latitude"), crs = 4326)

ggplot() +
  # Adds OpenStreetMap tiles automatically bounded by your data
  annotation_map_tile(type = "osm", zoom = 9) + 
  geom_sf(data = orcas_sf, aes(color = pods_or_ecotype), size = 1, alpha = 0.6) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    title = "Orca Encounter by Pods",
    color = "Pod / Ecotype"
  )
```


```{r}
#| echo: false
pods_vector <- orcas |> 
  group_by(pods_or_ecotype) |> 
  summarize(count = n()) |> 
  filter(count >= 2) |>
arrange(desc(count)) |>
head(4) |> 
pull(pods_or_ecotype)

# pods_vector
```


```{r}
#| eval: false
#| echo: false
library(ggplot2)
library(sf)
library(ggspatial)

# Convert coordinates to sf object
orcas_sf <- orcas %>%
  filter(!is.na(begin_longitude), !is.na(begin_latitude)) %>%
  st_as_sf(coords = c("begin_longitude", "begin_latitude"), crs = 4326)

ggplot() +
  # Adds OpenStreetMap tiles automatically bounded by your data
  annotation_map_tile(type = "osm", zoom = 9) + 
  geom_sf(
    data = filter(orcas_sf, pods_or_ecotype %in% pods_vector), 
    aes(color = pods_or_ecotype), 
    size = 1, 
    alpha = 0.6
  ) +
  # facet_wrap(~ pods_vector) +
  theme_minimal() +
  labs(
    title = "Orca Encounters by Most Frequent Pods",
    color = "Pod / Ecotype"
  )
```


```{r}
#| echo: false
library(dplyr)
library(ggplot2)
library(sf)
library(ggspatial)

# Convert coordinates to sf object
orcas_sf <- orcas %>%
  filter(!is.na(begin_longitude), !is.na(begin_latitude)) %>%
  st_as_sf(coords = c("begin_longitude", "begin_latitude"), crs = 4326)

# 1. Filter the dataset to keep ONLY the pods in pods_vector
orcas_filtered <- orcas_sf %>% 
  filter(pods_or_ecotype %in% pods_vector)

ggplot() +
  # Adds OpenStreetMap tiles automatically bounded by your data
  annotation_map_tile(type = "osm", zoom = 9) + 
  geom_sf(
    data = filter(orcas_filtered, pods_or_ecotype %in% pods_vector), 
    aes(color = pods_or_ecotype), 
    size = 1, 
    alpha = 0.6
  ) +
  facet_wrap(~ pods_or_ecotype, ncol = 2) + # Control column layout 
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 12, face = "bold") # Enlarge facet labels
  ) +
  labs(
    title = "Orca Encounters by Most Frequent Pods",
    color = "Pod / Ecotype"
  )
```

```{r}
#| eval: false
#| echo: false
# Preview the first few rows
head(orcas)

# Check summary statistics and missing values
glimpse(orcas)

# View unique values for pods or ecotypes observed
orcas %>% 
  count(pods_or_ecotype, sort = TRUE)

# Plot encounters over the years
orcas %>% 
  count(year) %>% 
  ggplot(aes(x = year, y = n)) +
  geom_col(fill = "steelblue") +
  theme_minimal() +
  labs(
    title = "CWR Orca Encounters per Year",
    x = "Year",
    y = "Number of Encounters"
  )
 
```